---
title: "667-Group4"
format: pdf
editor: visual
---

### Introduction


### GLM
GLM Method:

We modeled the probability of 30-day readmission using a logistic generalized linear model (GLM) with a logit link. The model takes the form:$log(\frac{P(readmit30=1)}{1-P(readmit30=1)}) = \beta_0 +\beta_1(insulin) + \beta_2(age) + ... + \beta_10(encounter  index)$ . Covariates included insulin prescription status, demographic variables (age, gender, race), encounter-level severity indicators (number of medications, time in hospital, admission type), prior healthcare utilization (emergency and inpatient visits), and a truncated encounter index reflecting recent encounter history. A marginal GLM was used as an initial approach because its coefficients yield population-average interpretations and provide a baseline for evaluating whether correlated-data models are necessary. Wald z-tests were used for inference on regression coefficients.

To evaluate the contribution of encounter history, we fit two versions of the GLM: one including the truncated encounter index and one excluding it. Comparing these models allowed us to assess whether encounter index added explanatory value and should be retained before proceeding to more advanced longitudinal modeling frameworks.

Regression coefficients were exponentiated to obtain odds ratios (ORs), and profile likelihood confidence intervals were computed using confint(). This allowed effects to be interpreted as multiplicative changes in the odds of readmission while maintaining likelihood-based inference.

Generalized variance inflation factors (GVIFs) were calculated to evaluate multicollinearity, particularly for categorical predictors with multiple degrees of freedom. This ensured stable coefficient estimation and interpretability.

We performed residual diagnostics, influence diagnostics, and goodness-of-fit assessments to evaluate whether the GLM adequately captured the relationship between predictors and 30-day readmission. Deviance and Pearson residuals were plotted against fitted probabilities to assess functional form, detect outlying patterns, and evaluate the appropriateness of the logit link. Overdispersion was examined by comparing the residual deviance to its degrees of freedom. Influence diagnostics were conducted using Cook’s distance to identify observations with disproportionate leverage on coefficient estimates. Model calibration was assessed using the Hosmer–Lemeshow test, and predictive discrimination was evaluated using the receiver operating characteristic (ROC) curve and its corresponding area under the curve (AUC). Together, these diagnostics provided a comprehensive assessment of model fit and predictive performance.

Because encounters were nested within patients, the independence assumption of a standard GLM may be violated. To explore potential within-patient dependence, we first generated encounter-level trajectory plots, which display how observed readmission proportions vary across encounter index and insulin categories. These plots provide an initial visual assessment of temporal patterns and possible effect modification over repeated encounters.

Next, we computed lag-1 Pearson residuals, plotting residuals at encounter t against residuals from encounter t−1. This diagnostic examines serial correlation among consecutive encounters; systematic dependence or diagonal structure in this plot would indicate violation of the independence assumption.

To evaluate whether the association between insulin status and readmission varied across encounter history, we fit a logistic GLM including insulin × encounter_index_trunc interaction terms. This model tested whether the effect of insulin depended on how many encounters a patient had previously accumulated. Model selection using AIC guided whether the interaction terms improved model specification.

Finally, to account for potential correlation directly in the inferential framework, we computed cluster-robust (sandwich) standard errors at the patient level. These robust SEs adjust for within-patient clustering while leaving coefficient estimates unchanged, allowing valid inference even when observations are correlated. Comparing model-based and cluster-robust standard errors helped assess the extent to which within-patient dependence affected uncertainty estimates and whether a marginal GLM remained adequate or if repeated-measures models (GEE or GLMM) might be required.

GLM Result:

```{r include=FALSE}
suppressPackageStartupMessages({
library(broom)
library(dplyr)
library(car)
library(ResourceSelection)
library(pROC)
library(knitr)
library(kableExtra)
library(ggplot2)
library(sandwich)
library(lmtest)
})
```

```{r include=FALSE}
dat <- read.csv("C:\\Users\\estel\\OneDrive\\Documents\\Desktop\\UNC\\BIOS667\\HW\\Final Project\\cleaned_diabetes_longitudinal_Nov29.csv")
dat$insulin <- relevel(factor(dat$insulin), ref = "No")
dat$age <- factor(dat$age)
dat$gender <- factor(dat$gender)
dat$race <- factor(dat$race)
dat$admission_type_id <- factor(dat$admission_type_id)

# Frequency table: how many patients have 2, 3, 4… encounters
enc_per_patient <- table(dat$patient_nbr)
table(enc_per_patient)

#Truncate encounter_index
dat$encounter_index_trunc <- pmin(dat$encounter_index, 5)
table(dat$encounter_index_trunc)

lapply(dat[, sapply(dat, is.factor)], levels)
```

```{r echo=FALSE}
#GLM with encounter_index_trunc
fit_glm <- glm(
  readmit30 ~ insulin + age + gender + race +
    num_medications + time_in_hospital +
    number_emergency + number_inpatient +
    admission_type_id + encounter_index_trunc,
  family = binomial(link = "logit"),
  data = dat
)

summary(fit_glm)

#GLM without encounter_index_trunc
fit_glm_2 <- glm(
  readmit30 ~ insulin + age + gender + race +
    num_medications + time_in_hospital +
    number_emergency + number_inpatient +
    admission_type_id,
  family = binomial(link = "logit"),
  data = dat
)
summary(fit_glm_2)
```

Here two logistic GLMs were fitted. One includes a truncated encounter index(capped at 5), and one without it. Including encounter_trunc improved model fit (AIC 4026 vs 4047) and reduced residual deviance (3992.2 vs 4015). The truncated index was highly significant(p<0.001). 

```{r or-table, echo=FALSE, message=FALSE}
OR_table <- exp(cbind(OR = coef(fit_glm), confint(fit_glm)))
pvals <- summary(fit_glm)$coef[, 4]

final_table <- cbind(OR_table, p.value = pvals)

kable(
  final_table,
  caption = "Odds Ratios (OR) with 95% Confidence Intervals for Logistic GLM",
  digits = 4
)
```

In the full GLM model, several predictors were significantly associated with 30-day readmission, including age 50–80, number of medications, prior emergency visits, prior inpatient visits, and the truncated encounter index.
Based on Table @ref(tab:or-table), patients aged 50–80 had about 20% lower odds of 30-day readmission compared with those ≤50 years (OR = 0.79, 95% CI: 0.61–0.98). This means that, holding all other variables constant, being in the 50–80 age group is associated with a meaningful reduction in readmission risk.
Each additional medication slightly raised the odds of readmission (OR = 1.01, 95% CI: 1.00–1.02), and greater numbers of emergency visits (OR = 1.07, 95% CI: 1.01–1.13) and inpatient visits (OR = 1.18, 95% CI: 1.10–1.27) were associated with higher odds. 
In contrast, the truncated encounter index showed a strong protective effect; each one-unit increase (up to the cap of 5) corresponded to roughly a 20% reduction in the odds of readmission (OR = 0.81, 95% CI: 0.73–0.88). Confidence intervals for the significant predictors did not cross 1, supporting their statistical significance.
InsulinSteady has a p-value slightly above 0.05, suggesting a possible increase in readmission risk, although its 95% confidence interval crossed 1 and therefore the effect was not statistically significant.

```{r vif-table, echo=FALSE}
# Diagnostics for GLM
# Multicollinearity (VIF)
vif_raw <- car::vif(fit_glm)
vif_tbl <- as.data.frame(vif_raw)
colnames(vif_tbl) <- c("GVIF", "Df", "GVIF^(1/(2×Df))")

vif_tbl <- vif_tbl %>%
  dplyr::mutate(across(everything(), ~ round(.x, 3)))

kable(
  vif_tbl,
  caption = "Variance Inflation Factors for Logistic GLM",
  align = "c"
)
```
Based on Table \@ref(tab:vif-table), all variance inflation factors were low (GVIF^(1/(2×Df)) ≈ 1–2), indicating no evidence of harmful multicollinearity among the predictors. Even the highest values, such asencounter_index_trunc (1.41) and number_inpatient (1.45), remain well below common concern thresholds (typically 2–5), suggesting that multicollinearity is minimal and unlikely to meaningfully bias the coefficient estimates.

```{r cooks-figure, echo=FALSE, fig.cap="Cook's Distance for Influential Observations"}
cooks <- cooks.distance(fit_glm)

plot(
  cooks,
  main = "Cook's Distance for Influential Observations",
  xlab = "Index",
  ylab = "Cook's Distance"
)

abline(h = 0.04, col = "red", lty = 2)

```
Based on Figure \@ref(fig:cooks-figure), all Cook’s Distance values are extremely small and far below commonly used influence thresholds. This indicates that no single observation had a notable impact on the fitted GLM.

```{r echo=FALSE}
#Check for Overdispersion
phi <- deviance(fit_glm) / df.residual(fit_glm)
phi
```
The dispersion estimate was φ = 0.97, which is very close to 1, indicating no evidence of overdispersion in the logistic GLM.

```{r devres-figure, echo=FALSE, fig.cap="Deviance Residuals vs Fitted Probabilities"}
plot(
  fitted(fit_glm),
  residuals(fit_glm, type = "deviance"),
  main = "Deviance Residuals vs Fitted Probabilities",
  xlab = "Fitted probabilities",
  ylab = "Deviance residuals",
  pch = 20, col = "#00000030"
)
abline(h = 0, col = "red")
```
The residual plot (Figure \@ref(fig:devres-figure)) shows a symmetric, funnel-shaped pattern typical of logistic regression and no systematic structure, indicating that the model fits reasonably well.

```{r pearson-figure, echo=FALSE, fig.cap="Pearson Residuals vs Fitted Probabilities"}
plot(
  fitted(fit_glm),
  residuals(fit_glm, type = "pearson"),
  main = "Pearson Residuals vs Fitted Probabilities",
  xlab = "Fitted probabilities",
  ylab = "Pearson residuals",
  pch = 20, col = "#00000030"
)
abline(h = 0, col = "red")
```
The Pearson residuals (Figure \@ref(fig:pearson-figure)) show the expected curved pattern for logistic regression and no systematic structure, indicating acceptable model fit.

```{r roc-figure, echo=FALSE, message=FALSE, fig.cap="ROC Curve for Logistic GLM"}
# Goodness-of-fit
y_glm  <- fit_glm$y
phat   <- fitted(fit_glm)
hoslem.test(y_glm, phat, g = 10)

# ROC Curve
roc_obj <- roc(y_glm, phat)
roc_obj
auc(roc_obj)

plot(
  roc_obj,
  main = "ROC Curve for Logistic GLM",   # no manual numbering here
  col = "blue"
)

```
The Hosmer–Lemeshow test gave a p-value of 0.66, indicates no evidence of poor calibration. The predicted and observed readmission rates are reasonably aligned. 
However, the AUC was 0.586, so the model’s ability to distinguish between readmitted and non-readmitted patients is only slightly better than chance.

```{r traj-figure, echo=FALSE, fig.cap="Trajectory of Readmission Rates Across Encounters, Stratified by Insulin Regimen"}
traj <- dat %>%
  dplyr::group_by(encounter_index_trunc, insulin) %>%
  dplyr::summarize(prop = mean(readmit30), .groups = "drop")

ggplot(traj, aes(x = encounter_index_trunc, y = prop, color = insulin)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Trajectory of Readmission Rates Across Encounters, Stratified by Insulin Regimen",
    x = "Encounter Index",
    y = "Readmission Proportion"
  )

```
The trajectory plot (Figure \@ref(fig:traj-figure))shows that the insulin groups follow clearly non-parallel paths across encounters, suggesting that the encounter effect is not the same for all groups.


```{r echo=FALSE}
#Interaction Check
##Fit a GLM with interaction: encounter_index * insulin
fit_interaction <- glm(
  readmit30 ~ insulin * encounter_index_trunc + age + gender + race +
    num_medications + time_in_hospital +
    number_emergency + number_inpatient + admission_type_id,
  family=binomial,
  data=dat
)

summary(fit_interaction)
```
Including the insulin × encounter_index_trunc interaction improved model fit (AIC decreased from 4026 to 4014), and several interaction terms were statistically significant. InsulinSteady×Encounter_index (p = 0.003) and InsulinUp×Encounter_index (p < 0.0001) showed strong evidence that the effect of encounter index differs by insulin group. This indicates that the association between encounter index and readmission is not constant across insulin categories.

The trajectory plot result is consistent with the significant interaction terms in the GLM. Together, they suggest that the slope linking encounter index to readmission probability varies by insulin status. This provides justification for modeling strategies that allow group-specific slopes or correlated repeated measures, rather than relying on a single common slope for all insulin groups.


```{r lag1-figure, echo=FALSE, fig.cap="Lag-1 Pearson Residual Plot"}
#Independence Diagnostics

##Lag-1 Pearson residual plot

#Extract Pearson residuals
pearson_resid <- residuals(fit_glm, type = "pearson")
pearson_resid <- pearson_resid[!is.na(pearson_resid)]

#Create lag-1 residual within each patient
lag1_resid <- dplyr::lag(pearson_resid, 1)

lag_df <- data.frame(
  r_t = pearson_resid[-1],
  r_t_minus1 = lag1_resid[-1]
)

#Lag-1 residual plot
ggplot(lag_df, aes(x = r_t_minus1, y = r_t)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Lag-1 Pearson Residual Plot",
    x = "Pearson Residual (t-1)",
    y = "Pearson Residual (t)"
  ) +
  theme_minimal()
```
The lag-1 Pearson residual plot (Figure \@ref(fig:lag1-figure)) shows clear clustering and diagonal patterns, indicating positive autocorrelation between residuals from consecutive encounters. This violates the independence assumption of the GLM and suggests that within-patient correlation is present, making methods such as GEE or GLMM more appropriate.

```{r rse-table, echo=FALSE}
#Robust SE Comparison (Sandwich)

# Model-based summary
model_sum <- summary(fit_glm)$coef

# Robust (clustered) standard errors
vcov_cl  <- vcovCL(fit_glm, cluster = ~ patient_nbr)
robust   <- coeftest(fit_glm, vcov = vcov_cl)

# Build comparison table
comp_table <- data.frame(
  Term       = rownames(model_sum),
  Estimate   = model_sum[, "Estimate"],
  SE_Model   = model_sum[, "Std. Error"],
  p_Model    = model_sum[, "Pr(>|z|)"],
  SE_Robust  = robust[, "Std. Error"],
  p_Robust   = robust[, "Pr(>|z|)"]
)

comp_table <- comp_table %>%
  mutate(across(-Term, ~ round(.x, 4)))

# Display table
rownames(comp_table) <- NULL

kable(
  comp_table,
  caption = "Model-Based vs. Cluster-Robust Standard Errors",
  align = "c"
) %>%
  kable_styling(full_width = FALSE)
```
Cluster-robust standard errors were computed to adjust for within-patient correlation across repeated encounters. While coefficient estimates were unchanged, several standard errors increased. Notably, the age 50–80 category was no longer statistically significant under the robust SEs (p = 0.066, from Table \@ref(tab:rse-table)), indicating that the naïve GLM slightly underestimated uncertainty. The main conclusions for number of medications, emergency visits, inpatient visits, and the encounter index remained robust.


### GEE

### GLMM

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(broom)
  library(broom.mixed)
  library(geepack)
  library(lme4)
  library(scales)
  library(MASS)      # for glm.nb
  library(performance)
  library(car)
  library(ggplot2)
  library(gridExtra) 
})
set.seed(667)
```

#### Exploratory plots

```{r}
data <- read_csv('../cleaned_diabetes_longitudinal_Nov29.csv')
glimpse(data)
data$patient_nbr <- as.factor(data$patient_nbr)
data$admission_type_id <- as.factor(data$admission_type_id)
data$race <- as.factor(data$race)
data$gender <- as.factor(data$gender)
data$age <- as.factor(data$age)
data$insulin <- as.factor(data$insulin)
data$readmit30 <- as.factor(data$readmit30)
data$age <- relevel(data$age, ref = "50-80")
data$insulin <- relevel(data$insulin, ref = "No")

data$encounter_index_trunc_c <-  data$encounter_index_trunc - mean(data$encounter_index_trunc)

data$time_in_hospital <- scale(data$time_in_hospital)
data$num_medications <- scale(data$num_medications)
data$number_emergency <- scale(data$number_emergency)
data$number_inpatient <- scale(data$number_inpatient)
```

Four numeric covariates were standardized to facilitate model convergence and improve the numerical stability of the optimization algorithm. Use centered time (encounter_index_trunc_c = encounter_index_trunc - mean(encounter_index_trunc)) for easier interpretation of main effects.

#### Fit random intercept model

We employed adaptive Gauss-Hermite quadrature with 10 integration points (nAGQ = 10) rather than the standard Laplace approximation (nAGQ = 1). For binary response data, the standard Laplace approximation is known to potentially underestimate the variance of random effects and introduce bias into the fixed effects coefficients. Increasing the number of quadrature points provides a more accurate approximation of the log-likelihood, thereby yielding more reliable parameter estimates.

```{r}
# Convergence using Laplace
m1 <- glmer(readmit30 ~ insulin * encounter_index_trunc_c + 
                      time_in_hospital + num_medications + 
                      number_emergency + number_inpatient + 
                      admission_type_id + age + gender + race +
                      (1 | patient_nbr), 
                    data = data, 
                    family = binomial(link = "logit"), nAGQ = 10,
                    control = glmerControl(optimizer = "bobyqa", 
                                           optCtrl = list(maxfun = 2e5)))
```

We need LRT to check the interaction term

```{r}
m2 <- glmer(readmit30 ~ insulin + encounter_index_trunc_c + 
            time_in_hospital + num_medications + 
            number_emergency + number_inpatient + 
            admission_type_id + age + gender + race +
            (1 | patient_nbr), 
            data = data, family = binomial, nAGQ = 10,
            control = glmerControl(optimizer = "bobyqa", 
                                   optCtrl = list(maxfun = 2e5)))
anova(m1, m2)
```

To assess whether the effect of time (encounter index) on readmission risk varies by insulin status, we compared the full model (including the interaction term) against a reduced model (main effects only) using a Likelihood Ratio Test (LRT).

The LRT indicated that the inclusion of the interaction term `insulin * encounter_index_trunc_c` resulted in a statistically significant improvement in model fit ($\chi^2(3) = 16.11$, $p < 0.005$). Furthermore, the Akaike Information Criterion (AIC) decreased substantially from 4019.0 in the reduced model to 4008.9 in the full model, confirming that the interaction model provides a superior fit to the data. This suggests that the trajectory of readmission risk over time is significantly different depending on the patient's insulin regimen.

#### Random Slopes Model

nAGQ = 10 can't be used in the random slope model, so we use nAGQ = 1 here.

```{r}
m3 <- glmer(readmit30 ~ insulin * encounter_index_trunc_c + 
            time_in_hospital + num_medications + 
            number_emergency + number_inpatient + 
            admission_type_id + age + gender + race +
            (1 + encounter_index_trunc_c | patient_nbr),       
            data = data, family = binomial, nAGQ = 1,
            control = glmerControl(optimizer = "bobyqa", 
                                   optCtrl = list(maxfun = 2e5)))

anova(m1, m3)
```

We evaluated the necessity of including random slopes for the time effect (`encounter_index_trunc_c`) to allow for inter-individual variation in the trajectory of readmission risk. A Likelihood Ratio Test comparing the random-slope model (`m3`) against the random-intercept-only model (`m1`) showed a statistically significant improvement in model fit ($\chi^2(2) = 14.42$, $p < 0.001$). Additionally, the random-slope model exhibited a lower AIC (3998.5 vs. 4008.9), indicating a better trade-off between goodness of fit and model complexity. Consequently, we selected the random-slope model (`m3`) as the final model. This suggests that not only do baseline risks vary across patients, but the rate of change in readmission risk over time also varies significantly between individuals.

```{r}
summary(m3)
# Check Convergence / ICC
m3@optinfo$conv$lme4
icc(m3)
```

------------------------------------------------------------------------

#### Interaction Effects

There was a significant interaction between insulin status and the timing of encounters. For the reference group (patients not on insulin), the probability of 30-day readmission showed a steep and significant decline over successive visits ($\beta_{time} = -0.462$, $p < 0.001$). However, this protective temporal trend was significantly attenuated for patients on steady insulin ($\beta_{interaction} = 0.232$, $p = 0.014$) and further diminished for patients whose insulin dosage was increased ($\beta_{interaction} = 0.329$, $p = 0.004$). This indicates that while Insulin Up patients did experience a marginal reduction in risk over time, their **rate of recovery was significantly slower and flatter** compared to patients not requiring insulin therapy.

#### ICC and Variance

The Adjusted Intraclass Correlation Coefficient (ICC) was 0.113, indicating that approximately 11.3% of the latent variation in readmission risk is attributable to unobserved heterogeneity between patients (such as unique genetics, chronic frailty, or compliance behavior). The remaining variation is driven by within-patient time-varying factors or random noise. An ICC exceeding 10% confirms substantial clustering within subjects, validating the necessity of the mixed-effects framework to account for non-independence in the data.

#### Covariance Matrix

The model's direct Maximum Likelihood Estimation (MLE) of the covariance matrix yielded a correlation coefficient of only $\rho = -0.04$. This parameter represents the unbiased population-level relationship. A correlation of -0.04 is statistically negligible, indicating that in the true underlying population, a patient's baseline risk level is effectively independent of their rate of change over time.

#### Diagnostic

```{r}
rand_effects <- ranef(m3)$patient_nbr

colnames(rand_effects) <- c("Intercept", "Slope")

p_qq_int <- ggplot(rand_effects, aes(sample = Intercept)) +
  stat_qq(color = "blue", alpha = 0.5) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Q-Q Plot: Random Intercepts",
       subtitle = "Normality of Baseline Risk",
       x = "Theoretical", y = "Sample") +
  theme_minimal()

p_qq_slope <- ggplot(rand_effects, aes(sample = Slope)) +
  stat_qq(color = "darkgreen", alpha = 0.5) +
  stat_qq_line(color = "red", linetype = "dashed") +
  labs(title = "Q-Q Plot: Random Slopes",
       subtitle = "Normality of Time Effect",
       x = "Theoretical", y = "Sample") +
  theme_minimal()

p_scatter <- ggplot(rand_effects, aes(x = Intercept, y = Slope)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  geom_vline(xintercept = 0, linetype="dashed", alpha=0.3) +
  geom_hline(yintercept = 0, linetype="dashed", alpha=0.3) +
  labs(title = "Intercept vs. Slope Correlation",
       subtitle = paste0("Cor: ", round(cor(rand_effects$Intercept, rand_effects$Slope), 3)),
       x = "Random Intercept (Baseline Risk)",
       y = "Random Slope (Rate of Change)") +
  theme_minimal()

grid.arrange(p_qq_int, p_qq_slope, p_scatter, ncol = 2)
```

**Evaluation of Random Effects Structure:** The diagnostic plots for the random effects structure (Model `m3`) revealed deviations from ideal normality for both random intercepts and random slopes. The Q-Q plots exhibit a "stepped" pattern, which is a characteristic artifact in generalized linear mixed models (GLMMs) when the outcome is binary and cluster sizes are small (i.e., limited visits per patient).

Despite these deviations, two key insights emerged: 1. Correlation of Effects: We observed a moderate negative correlation ($\rho = -0.291$) between the random intercepts and slopes. This suggests a regression to the mean dynamic: patients with higher baseline risks (higher intercepts) tend to experience a sharper decline in risk over time (more negative slopes), whereas lower-risk patients exhibit more stable trajectories. 2. Model Robustness: Although the random effects distributions are not perfectly normal, previous simulation studies in GLMMs have demonstrated that fixed effect estimates remain robust to misspecification of the random effects distribution, especially given our large sample size. Therefore, retaining the random-slope structure provides a more realistic representation of patient heterogeneity without compromising the validity of the fixed effect inferences.

The discrepancy between the visual estimate (-0.291) and the model estimate (-0.04) is explained by the **'Shrinkage Effect'** common in GLMMs. Because many patients have few visits, the model 'shrinks' their uncertain individual estimates toward the population mean, which mathematically forces a spurious negative correlation in the scatterplot. **Therefore, we rely on the model output (-0.04) as the valid inference.** Our final clinical conclusion is that a patient's starting frailty (intercept) does not predict how fast they will stabilize (slope); these are two distinct biological characteristics.

```{r}
isSingular(m3)
vif(m3)
```

We assessed multicollinearity using the Generalized Variance Inflation Factor (GVIF). All predictors showed low GVIF values, falling well below the conservative threshold of 5.

#### Simulation

```{r}
library(ggplot2)
library(dplyr)

vc <- attr(VarCorr(m3)$patient_nbr, "stddev")
sigma_int <- vc["(Intercept)"]
sigma_slp <- vc["encounter_index_trunc_c"]

beta_int   <- fixef(m3)["(Intercept)"]
beta_time  <- fixef(m3)["encounter_index_trunc_c"]
beta_ins   <- fixef(m3)["insulinSteady"]
beta_inter <- fixef(m3)["insulinSteady:encounter_index_trunc_c"]

avg_fixed <- mean(predict(m3, type="link", re.form=NA))

visits <- seq(1, 6, length.out = 100)
visits_c <- visits - mean(data$encounter_index_trunc_c)

scenarios <- expand.grid(
  Visit = visits,
  Baseline_Type = c("High Baseline", "Avg Baseline", "Low Baseline"),
  Recovery_Speed = c("Slow Recovery (Flat)", "Fast Recovery (Steep)")
)

scenarios <- scenarios %>%
  mutate(
    z_int = case_when(
      Baseline_Type == "High Baseline" ~ 1.96,
      Baseline_Type == "Avg Baseline"  ~ 0,
      Baseline_Type == "Low Baseline"  ~ -1.96
    ),
    z_slp = case_when(
      Recovery_Speed == "Slow Recovery (Flat)"  ~ 1.96,
      Recovery_Speed == "Fast Recovery (Steep)" ~ -1.96
    )
  )

# LogOdds = Fixed_Part + (z_int * sigma_int) + (Fixed_Slope + z_slp * sigma_slp) * Time
scenarios$LogOdds <- avg_fixed + 
                     (scenarios$z_int * sigma_int) + 
                     ((beta_time + beta_inter + scenarios$z_slp * sigma_slp) * visits_c) 
scenarios <- scenarios %>%
  mutate(
    visits_centered = Visit - mean(data$encounter_index_trunc_c),
    LogOdds = avg_fixed + 
              (z_int * sigma_int) + 
              (beta_time + beta_inter + z_slp * sigma_slp) * visits_centered,
    Prob = plogis(LogOdds)
  )

ggplot(scenarios, aes(x = Visit, y = Prob, color = Baseline_Type, linetype = Recovery_Speed)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("grey60", "red", "blue")) + # Avg, High, Low
  scale_linetype_manual(values = c("solid", "dashed")) +    # Fast, Slow
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  labs(
    title = "Patient Heterogeneity: Baseline vs. Recovery Speed",
    subtitle = "Decoupling 'Where they start' from 'How fast they improve' (Model m3)",
    x = "Encounter Number",
    y = "Predicted Probability of Readmission",
    color = "Starting Risk (Intercept)",
    linetype = "Response Rate (Slope)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.key.width = unit(2, "cm"),
    plot.title = element_text(face = "bold")
  )
```

Our Random Slope model (m3) reveals a more nuanced reality because the correlation between baseline and slope is near zero.

In this simulation, we visualize 6 scenarios: \> **The Color (Baseline Risk):** Shows where a patient starts. Red is high risk, Blue is low risk. \> **The Linetype (Recovery Speed):** Shows how they respond to repeated visits. \> **Solid Line (Fast Recovery):** These patients respond well; their risk drops steeply. \> **Dashed Line (Slow Recovery):** These patients are 'resistant' to improvement; their risk stays flat.

Crucially, look at the Red Solid Line vs. the Red Dashed Line. Both start as 'High Risk' patients. But one responds rapidly to care (Solid), while the other remains chronically ill (Dashed). Identifying which 'High Risk' patient we are dealing with is critical for personalized resource allocation."

```{r}
preds <- augment(m3, data = data, type.predict = "response")

set.seed(42)
sample_subjects <- sample(unique(data$patient_nbr), 9)

preds %>%
  filter(patient_nbr %in% sample_subjects) %>%
  ggplot(aes(x = encounter_index_trunc_c, y = .fitted, color = insulin, group = patient_nbr)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_jitter(aes(y = as.numeric(as.character(readmit30))), 
              height = 0.02, width = 0.05, shape = 21, fill = "white", size = 2.5, stroke = 1) +
  
  labs(
    y = "Predicted Probability of Readmission",
    x = "Encounter Number",
    title = "Subject-Specific Predicted Trajectories (Sample n=9)",
    subtitle = "Lines = GLMM fitted values (BLUPs); Points = Observed Outcomes (0=No, 1=Yes)",
    color = "Insulin Status"
  ) +

  scale_y_continuous(labels = scales::percent, limits = c(-0.05, 1.05), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = 1:6) +
  scale_color_brewer(palette = "Set1") + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )
```

```{r}
sigma2_b <- VarCorr(m3)$patient_nbr[1,1]
print(paste0("Random Intercept Variance (sigma^2_b): ", round(sigma2_b, 4)))

# Extract treatment effect
cond_est <- tidy(m3, effects = "fixed", conf.int = TRUE) %>%
  dplyr::select(term, estimate, std.error, conf.low, conf.high)

# Apply Zeger attenuation formula
zeger_factor <- sqrt(1 + 0.346 * sigma2_b)
attenuation_pct <- (1 - 1/zeger_factor) * 100

print(paste0("Zeger Attenuation Factor: ", round(zeger_factor, 4)))
print(paste0("Attenuation Percentage: ", round(attenuation_pct, 2), "%"))

zeger_table <- cond_est %>%
  mutate(
    # Conditional Results
    Cond_Beta = estimate,
    Cond_OR = exp(estimate),
    # Zeger Approximate Marginal Results
    Marg_Beta_Approx = estimate / zeger_factor,
    Marg_OR_Approx = exp(estimate / zeger_factor)
  ) %>%
  dplyr::select(term, Cond_Beta, Cond_OR, Marg_Beta_Approx, Marg_OR_Approx) %>%
  filter(term != "(Intercept)") # Focus on effects
knitr::kable(zeger_table, digits = 3, caption = "Conditional (GLMM) vs. Approximate Marginal (Zeger) Effects")
```

------------------------------------------------------------------------

#### Results: Analysis of 30-Day Readmission Risk (Random-Slope GLMM)

##### 1. Model Selection and Specification

We employed a Generalized Linear Mixed Model (GLMM) to assess predictors of 30-day readmission. While the initial model included only random intercepts, a Likelihood Ratio Test (LRT) indicated that adding random slopes for the time effect significantly improved model fit ($\chi^2 = 14.42, p < 0.001$). Consequently, the final model (`m3`) included both random intercepts and random slopes, allowing each patient to have a unique baseline risk and a unique trajectory of risk evolution over time.

##### 2. Random Effects and Subject Heterogeneity

The final model revealed substantial patient heterogeneity. The variance of the random intercept was **0.175** ($\text{SD} = 0.418$), indicating significant differences in initial frailty across patients. The variance of the random slope was **0.165** ($\text{SD} = 0.406$), confirming that patients recover (or deteriorate) at significantly different rates. The correlation between random intercepts and slopes was negligible ($\rho = -0.04$), suggesting that a patient's baseline risk is largely independent of their rate of recovery. The Adjusted Intraclass Correlation Coefficient (ICC) was **0.113**, implying that approximately **11.3%** of the variation in readmission risk is attributable to unobserved subject-specific characteristics.

##### 3. Fixed Effects: The Interaction of Insulin and Time

The analysis demonstrated a significant interaction between insulin status and the timing of encounters. Protective Trend in Reference Group: For patients not on insulin, the probability of readmission decreased steeply with successive visits ($\text{Cond. OR} = 0.630$, $\beta = -0.462, p < 0.001$), indicating a strong 'learning' or stabilization effect. Attenuation in Insulin Groups:\*\* This protective temporal trend was significantly blunted for patients on insulin. The interaction term for the Insulin Steady group was significant ($\beta_{interaction} = 0.232, p = 0.014$), and even more pronounced for the Insulin Up group ($\beta_{interaction} = 0.329, p = 0.004$). While 'Insulin Up' patients still experienced a slight decrease in risk over time (Net slope $\approx -0.13$), their rate of improvement was significantly slower compared to the reference group.

##### 4. Conditional vs. Approximate Marginal Effects (Zeger Method)

Since GLMM parameters represent subject-specific (conditional) effects, we applied Zeger's method to approximate population-averaged (marginal) effects.The Zeger attenuation factor was **1.03**, resulting in a modest **2.89% reduction** in effect sizes when scaling from the individual to the population level. The Conditional ORs and Marginal ORs remained highly consistent (e.g., Insulin Up Cond. OR = 1.129 vs. Marg. OR = 1.125). This suggests that despite patient heterogeneity, the observed risk factors apply robustly to the population average.

##### 5. Other Clinical Predictors

Controlling for the time-insulin interaction, other clinical factors showed the following associations: Prior inpatient visits remained a significant predictor, with each additional visit increasing the odds of readmission by 16.4% ($\text{Cond. OR} = 1.164, p = 0.019$).The number of medications showed a borderline significant association ($\text{Cond. OR} = 1.100, p = 0.051$), serving as a likely proxy for comorbidity complexity. Age, gender, and race were not statistically significant in this adjusted longitudinal framework.

------------------------------------------------------------------------

#### Summary Conclusion

In conclusion, the **Random Slope model** provides superior fit, revealing that patients differ not only in their baseline risk but also in their recovery trajectories. While the "No Insulin" group exhibits a rapid decline in readmission risk over repeated encounters, patients requiring insulin dose escalation ("Insulin Up") show a significantly **flatter recovery curve**. This suggests that insulin intensification serves as a marker for a more resistant, high-risk phenotype that benefits less from the standard "stabilization over time" effect.

### Comparison

### Discussion

### Conclusion
